---
title: "Body Fat Prediction Model — Regression Analysis for Health & Wellness Applications"
author: "Harris K."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  pdf_document:
    toc: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options("scipen" = 10)

loadLib <- function(libName) {
  if (!require(libName, character.only = TRUE)) {
    install.packages(libName)
    require(libName, character.only = TRUE)
  }
}

libs <- c(
  "faraway", "glmnet", "MASS", "olsrr", "dplyr",
  "reshape2", "knitr", "caret", "ggplot2", "scales", "lmtest"
)
for (lib in libs) loadLib(lib)

# Helper — prints model equation from a fitted lm object
dispRegFunc <- function(reg) {
  coefs <- reg$coefficients
  b0 <- coefs[1]
  n <- length(coefs)
  eq <- paste0("Y = ", round(b0, 4))
  for (i in 2:n) {
    eq <- paste0(eq, " + ", round(coefs[i], 4), "*", names(coefs)[i])
  }
  cat(eq, "\n")
}
```

---

## 1. Project Overview

### 1.1 Objective

This analysis builds a regression-based body fat prediction model intended for integration
into a health and wellness application. The model estimates body fat percentage from
physical measurements that users can self-report during onboarding — no clinical equipment
required.

The body fat estimate feeds directly into a personalized calorie calculation engine:

- **Weight loss:** Calorie deficit relative to TDEE based on lean body mass
- **Weight maintenance:** Calorie target matched to TDEE
- **Weight gain:** Calorie surplus with lean mass preservation targets

### 1.2 Why Not BMI?

BMI is the default proxy for body composition in most consumer apps, but it is a crude
measure. A muscular user and an overweight user can have the same BMI while having very
different body fat percentages — and therefore very different calorie needs. A direct
body fat model improves recommendation quality significantly for a large portion of users.

### 1.3 Modeling Strategy

Five regression approaches are tested and compared:

| Model | Key Property |
|---|---|
| Linear Regression | Baseline — all predictors included |
| Stepwise Selection | Automatic variable selection, both directions |
| Ridge Regression | L2 regularization — shrinks all coefficients |
| Lasso Regression | L1 regularization — performs variable selection |
| Elastic Net | Blends Ridge and Lasso penalties |

A Huber robust regression is also run to assess how much outlier observations affect
model stability — a practical concern for a model that will see diverse user body types.

All models are trained on 70% of the data and evaluated on a held-out 30% test set.

---

## 2. Data Preparation

### 2.1 Dataset

```{r load-data}
data(fat, package = "faraway")

# Remove variables that are derived from the target and
# would not be available at prediction time in a live app
fat.data <- fat[, !(names(fat) %in% c("brozek", "density"))]

cat("Observations:", nrow(fat.data), "\n")
cat("Predictors:  ", ncol(fat.data) - 1, "\n")
cat("Response:     siri (body fat %)\n")
```

The dataset contains 252 adult male subjects. The response variable `siri` is body fat
percentage measured via underwater weighing — the gold standard clinical method. Predictor
variables include age, weight, height, BMI proxy (`adipos`), and 10 circumference
measurements.

### 2.2 Train / Test Split

```{r split}
set.seed(1023)
train.idx  <- sample(1:nrow(fat.data), size = floor(0.70 * nrow(fat.data)))
train.data <- fat.data[train.idx, ]
test.data  <- fat.data[-train.idx, ]

cat("Train set:", nrow(train.data), "observations (70%)\n")
cat("Test set: ", nrow(test.data),  "observations (30%)\n")
```

### 2.3 Exploratory Analysis

```{r eda-boxplot}
boxplot(fat.data$siri,
        main  = "Distribution of Body Fat Percentage (siri)",
        ylab  = "Body Fat %",
        col   = "steelblue",
        notch = FALSE)
```

```{r eda-correlation}
# Correlation between all predictors and the response
cor.with.siri <- sort(cor(fat.data)[, "siri"], decreasing = TRUE)
cor.df <- data.frame(
  Predictor   = names(cor.with.siri),
  Correlation = round(cor.with.siri, 3)
)
knitr::kable(cor.df, caption = "Correlation of Each Predictor with Body Fat %")
```

```{r eda-pairs, fig.height=8}
# Scatterplot matrix for the strongest predictors
strong.predictors <- c("siri", "abdom", "adipos", "chest", "hip", "weight")
pairs(fat.data[, strong.predictors],
      main = "Scatterplot Matrix — Top Predictors vs Body Fat %",
      pch  = 19, col = "steelblue", cex = 0.5)
```

Abdomen circumference has the strongest correlation with body fat (r = 0.81), followed
by adiposity index, chest, and hip. Body fat distribution is roughly normal with a slight
right skew and a few high-end outliers.

---

## 3. Model Building

### 3.1 Linear Regression — Full Model

```{r linear}
fat.lm <- lm(siri ~ ., data = train.data)
summary(fat.lm)
dispRegFunc(fat.lm)
```

```{r linear-diagnostics, fig.height=6}
par(mfrow = c(2, 2))
plot(fat.lm, main = "Linear Regression — Diagnostic Plots")
```

```{r linear-bp}
olsrr::ols_test_breusch_pagan(fat.lm)
```

```{r linear-vif}
vif.df <- data.frame(VIF = round(vif(fat.lm), 2))
vif.df$Predictor <- rownames(vif.df)
rownames(vif.df) <- NULL
knitr::kable(
  vif.df[order(-vif.df$VIF), c("Predictor", "VIF")],
  caption = "Variance Inflation Factors — Linear Model"
)
```

The full linear model achieves R² of 97.22% on training data. However, only 5 of 15
predictors are individually significant — a strong signal of multicollinearity.
VIF analysis confirms this: `weight` has a VIF of 199.54, and 6 predictors exceed
the threshold of 10. The Breusch-Pagan test result suggests checking for heteroscedasticity
before relying on standard errors.

**Takeaway for the app:** The full linear model fits training data very well but is likely
overfitting — too many correlated predictors are pulling in different directions.

---

### 3.2 Stepwise Selection

```{r stepwise}
set.seed(1023)
fat.sw       <- olsrr::ols_step_both_p(fat.lm, pent = 0.35, prem = 0.05, details = FALSE)
fat.sw.final <- fat.sw$model
summary(fat.sw.final)
dispRegFunc(fat.sw.final)
plot(fat.sw)
```

Stepwise selection reduced the model from 15 to 9 predictors while maintaining nearly
identical training R² (97.21%). Removed predictors: `abdom`, `hip`, `neck`, `knee`,
`ankle`, `biceps`. The retained variables represent a leaner, more interpretable model.

**Takeaway for the app:** The 9-predictor stepwise model is more practical for onboarding
— fewer measurements to ask users for.

---

### 3.3 Ridge Regression

```{r ridge-vif}
# Multicollinearity confirmed — Ridge is appropriate
olsrr::ols_vif_tol(fat.lm)
```

```{r ridge-corr}
# Pairs of predictors with correlation above 0.80
fat.cor <- data.frame(cor(train.data))
fat.cor$P1 <- colnames(fat.cor)
fat.cor.long <- fat.cor %>%
  reshape2::melt(id.vars = "P1", variable.name = "P2", value.name = "r") %>%
  dplyr::filter(r >= 0.80 & P1 != P2) %>%
  dplyr::arrange(desc(r)) %>%
  dplyr::filter(row_number() %% 2 == 1)
knitr::kable(fat.cor.long, caption = "Highly Correlated Predictor Pairs (r >= 0.80)")
```

```{r ridge-model}
set.seed(1023)
x.train <- data.matrix(dplyr::select(train.data, -siri))
y.train <- train.data$siri

fat.ridge.cv    <- glmnet::cv.glmnet(x.train, y.train, alpha = 0,
                                      nlambda = 100, lambda.min.ratio = 0.0001)
best.lambda.ridge <- fat.ridge.cv$lambda.min
plot(fat.ridge.cv)
cat("Optimal lambda (Ridge):", round(best.lambda.ridge, 3), "\n")

ridge.coefs <- coef(fat.ridge.cv, s = best.lambda.ridge)
knitr::kable(
  data.frame(
    Predictor = ridge.coefs@Dimnames[[1]],
    Coefficient = round(ridge.coefs@x, 4)
  ),
  caption = "Ridge Regression Coefficients at Optimal Lambda"
)
```

Ridge addresses multicollinearity by applying an L2 penalty that shrinks all coefficients
toward zero proportionally. The optimal lambda is 0.626. All 15 predictors are retained
but their coefficients are redistributed — no single measurement dominates the prediction.

Notable: the `wrist` coefficient flipped sign from positive (+0.263 in OLS) to negative
(-0.352 in Ridge), because Ridge redistributed wrist's effect among correlated measurements
for a more stable fit.

**Takeaway for the app:** Ridge is the most numerically stable model for users with
unusual body proportions, where correlated measurements might otherwise push OLS predictions
to extremes.

---

### 3.4 Lasso Regression

```{r lasso}
set.seed(1023)
fat.lasso.cv    <- glmnet::cv.glmnet(x.train, y.train, alpha = 1,
                                      nlambda = 100, lambda.min.ratio = 0.0001)
best.lambda.lasso <- fat.lasso.cv$lambda.min
plot(fat.lasso.cv)
cat("Optimal lambda (Lasso):", round(best.lambda.lasso, 3), "\n")

lasso.coefs <- coef(fat.lasso.cv, s = "lambda.min")
knitr::kable(
  data.frame(
    Predictor   = lasso.coefs@Dimnames[[1]],
    Coefficient = round(lasso.coefs@x, 4)
  ),
  caption = "Lasso Coefficients at Optimal Lambda"
)
```

Lasso uses a smaller optimal lambda (0.015) than Ridge, meaning less penalty is needed
to find a good model. Only one predictor (`hip`) is dropped entirely — its coefficient
set to exactly zero. The `free` variable (fat-free body weight) remains the strongest
predictor at -0.587, consistent with the linear model.

**Takeaway for the app:** Lasso confirms that nearly all body measurements carry
independent predictive value. `hip` can be dropped from the user onboarding form
without meaningful accuracy loss.

---

### 3.5 Elastic Net

```{r elastic-net}
set.seed(1023)
fat.enet.cv    <- glmnet::cv.glmnet(x.train, y.train, alpha = 0.5,
                                     nlambda = 100, lambda.min.ratio = 0.0001)
best.lambda.enet <- fat.enet.cv$lambda.min
cat("Optimal lambda (Elastic Net):", round(best.lambda.enet, 3), "\n")

enet.coefs <- coef(fat.enet.cv, s = best.lambda.enet)
knitr::kable(
  data.frame(
    Predictor   = enet.coefs@Dimnames[[1]],
    Coefficient = round(enet.coefs@x, 4)
  ),
  caption = "Elastic Net Coefficients at Optimal Lambda"
)
```

Elastic Net (alpha = 0.5) blends both penalties. Its optimal lambda of 0.039 sits between
Lasso (0.015) and Ridge (0.626) as expected. Like Lasso, it removes only `hip`. All
remaining coefficients are consistent with the other regularized models — a good sign of
stable feature importance across methods.

---

## 4. Outlier Analysis — Robust Regression

```{r robust}
fat.huber <- MASS::rlm(siri ~ ., psi = psi.huber, data = fat.data)
summary(fat.huber)
dispRegFunc(fat.huber)
```

```{r robust-diagnostics}
# Cook's distance — identifies high-influence observations
olsrr::ols_plot_cooksd_chart(lm(siri ~ ., data = fat.data))

# Residual-leverage plot
olsrr::ols_plot_resid_lev(lm(siri ~ ., data = fat.data))
```

```{r robust-weights}
fat.data$huber.weight <- fat.huber$w

fat.data %>%
  dplyr::mutate(Observation = row_number()) %>%
  dplyr::arrange(huber.weight) %>%
  dplyr::filter(row_number() <= 5) %>%
  dplyr::select(Observation, siri, weight, huber.weight) %>%
  knitr::kable(caption = "5 Observations with Lowest Huber Weights (Most Unusual)")
```

```{r robust-scatter}
ggplot2::ggplot(fat.data, ggplot2::aes(x = height, y = weight)) +
  ggplot2::geom_point(color = "steelblue", alpha = 0.7) +
  ggplot2::geom_text(
    label = rownames(fat.data),
    nudge_x = 0.3, nudge_y = 0.3,
    check_overlap = TRUE, size = 3
  ) +
  ggplot2::labs(
    title = "Weight vs Height — Outlier Detection",
    x = "Height (inches)", y = "Weight (lbs)"
  ) +
  ggplot2::theme_minimal()
```

Two subjects received the lowest Huber weights, meaning the model downweighted their
influence during fitting:

- **Observation 39:** Weight of 363 lbs, well outside the typical range. Body fat
  measured at 35.2% — much higher than the model would predict from circumference
  measurements alone. In a live app, this user should be flagged for manual review
  rather than receiving a raw model output.

- **Observation 221:** Normal body dimensions but unusually low body fat (12.4%),
  suggesting either exceptional fitness or a measurement anomaly.

**Takeaway for the app:** A production deployment should include a confidence interval
on each prediction. Users whose measurements fall more than 2 standard deviations from
the training distribution should receive a disclaimer alongside their estimate.

---

## 5. Model Comparison

### 5.1 Test Set Performance

```{r test-comparison}
set.seed(1023)
x.test <- data.matrix(dplyr::select(test.data, -siri))

test.data$pred.lm    <- predict(fat.lm, test.data)
test.data$pred.sw    <- predict(fat.sw.final, test.data)
test.data$pred.ridge <- predict(fat.ridge.cv, s = best.lambda.ridge, newx = x.test)[, 1]
test.data$pred.lasso <- predict(fat.lasso.cv, s = best.lambda.lasso, newx = x.test)[, 1]
test.data$pred.enet  <- predict(fat.enet.cv,  s = best.lambda.enet,  newx = x.test)[, 1]

sst <- sum((test.data$siri - mean(test.data$siri))^2)
n   <- nrow(test.data)

results <- data.frame(
  Model = c("Linear Regression", "Stepwise", "Ridge", "Lasso", "Elastic Net"),
  Train_R2 = c(0.9722, 0.9721, NA, NA, NA),
  Test_R2  = round(c(
    1 - sum((test.data$siri - test.data$pred.lm)^2)    / sst,
    1 - sum((test.data$siri - test.data$pred.sw)^2)    / sst,
    1 - sum((test.data$siri - test.data$pred.ridge)^2) / sst,
    1 - sum((test.data$siri - test.data$pred.lasso)^2) / sst,
    1 - sum((test.data$siri - test.data$pred.enet)^2)  / sst
  ), 4),
  Test_MSE = round(c(
    sum((test.data$siri - test.data$pred.lm)^2)    / n,
    sum((test.data$siri - test.data$pred.sw)^2)    / n,
    sum((test.data$siri - test.data$pred.ridge)^2) / n,
    sum((test.data$siri - test.data$pred.lasso)^2) / n,
    sum((test.data$siri - test.data$pred.enet)^2)  / n
  ), 4)
) %>% dplyr::arrange(desc(Test_R2))

knitr::kable(results, caption = "Model Performance on Held-Out Test Data")
```

### 5.2 Results Summary

```{r results-chart}
ggplot2::ggplot(results, ggplot2::aes(x = reorder(Model, Test_R2), y = Test_R2)) +
  ggplot2::geom_col(fill = "steelblue", width = 0.6) +
  ggplot2::geom_text(ggplot2::aes(label = round(Test_R2, 3)),
                     hjust = -0.1, size = 3.5) +
  ggplot2::coord_flip() +
  ggplot2::ylim(0, 1) +
  ggplot2::labs(
    title = "Test Set R² by Model",
    x     = NULL,
    y     = "R² (Test Data)"
  ) +
  ggplot2::theme_minimal()
```

Ridge regression achieves the highest test R² (0.876) and lowest MSE (9.67). Linear and
Stepwise both trained at 97% R² but dropped significantly on test data — a clear sign
of overfitting. Regularized models generalize far better to new users.

---

## 6. Production Recommendation

### 6.1 Chosen Model: Ridge Regression

Based on test set performance, Ridge regression is recommended for production deployment.

| Metric | Value |
|---|---|
| Test R² | 0.876 |
| Test MSE | 9.67 |
| Predictors Required | 15 |
| Handles Multicollinearity | Yes |
| Overfitting Risk | Low |

### 6.2 Downstream Calorie Calculation

Once body fat percentage is estimated, calorie targets can be calculated as follows:

```
Lean Body Mass (lbs) = Weight × (1 - BodyFat% / 100)
BMR = 370 + (21.6 × LBM in kg)                   # Katch-McArdle formula
TDEE = BMR × Activity Multiplier

Weight Loss:       Calorie Target = TDEE - 500     # ~0.5 lb/week deficit
Maintenance:       Calorie Target = TDEE
Weight Gain:       Calorie Target = TDEE + 300     # lean bulk surplus
```

### 6.3 Edge Case Handling

- Users whose body measurements fall outside the training distribution should receive
  a confidence interval alongside their body fat estimate
- Observation flagging logic should be implemented before returning predictions to the UI
- Model should be retrained periodically as the user base grows and more measurement
  data becomes available

### 6.4 Minimum Required Inputs for Prediction

Based on Lasso and Elastic Net results, the minimum effective input set is:

`age`, `weight`, `height`, `adipos`, `free`, `neck`, `chest`, `thigh`,
`forearm`, `wrist`, `knee`, `ankle`, `biceps`

(`hip` can be excluded — it was the only predictor removed by regularization.)

---

## 7. Conclusion

This analysis demonstrates that body fat percentage can be predicted with high accuracy
(test R² = 0.876) from self-reportable physical measurements using Ridge regression.
The model is robust, stable across unseen data, and practical for integration into a
consumer health app.

Replacing BMI-based calorie estimates with this body composition model would meaningfully
improve recommendation quality — particularly for muscular users, users carrying weight
in specific body regions, and users at the extremes of the weight distribution.

The model is ready for integration into a calorie calculation pipeline with the
downstream formula described in Section 6.2.
```
